---
output:
  html_document: default
  pdf_document: default
---
```{r setup, include = F}
library(tidyverse)
source("../R/helpers.R")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r, include = F, cache = T}
source("../R/main-data.R")
```

## Re: Grant - power calculations

---

**Summary:** Using preliminary RNA-seq data from four donors to estimate the variability of IRAK3 expression, we estimated the effect size detectable with 80% power for the proposed sample size of 10 normal and 10 severe asthma donors. Assuming comparable variability in the future cohort, the study would have 80% power to detect approximately 1.2–1.6-fold differences in IRAK3 expression between groups, depending on treatment and cell type.

These estimates are based on preliminary data from only four donors and therefore should be considered approximate; they assume that between-donor variability in the planned cohort is comparable to that observed in the preliminary dataset.

---

#### Example calculation

Pull IRAK3 data for HBEs at NS:

```{r}
df = primary_tpm %>%
  filter(Gene == "IRAK3", celltype == "HBE") %>%
  filter(treatment == "NS")

df
```

Calculate standard deviation of log2tpm values:

```{r}
sd_estimate <- sd(df$log2tpm)

sd_estimate
```

Given a two-sample design, N=10, alpha=0.05, and standard deviation from above, what is the power to detect a 2-fold (delta=1 in log2 space) change between normal and asthmatic?

```{r}
power.t.test(
  n = 10,
  delta = 1,
  sd = sd_estimate,
  sig.level = 0.05,
  type = "two.sample",
  alternative = "two.sided"
)
```

So, 99.9% probability of correctly detecting a doubling of IRAK3 expression between groups. This does not mean that the study has 99.9% power overall.

#### Preliminary power analysis for between-group differences in IRAK3 expression

We can scale out the above calculation to IRAK3 in all primary cell types, across a range of gene expression changes:

```{r}
power_results <- primary_tpm %>%
  filter(Gene == "IRAK3", celltype %in% c("ALI", "HBE")) %>%

  group_by(treatment, celltype) %>%
  summarise(
    sd_estimate = sd(log2tpm, na.rm = TRUE),
    .groups = "drop"
  ) %>%

  group_by(treatment, celltype) %>%
  mutate(
    `power_0.25` = power.t.test(
      n = 10,
      delta = 0.25,
      sd = sd_estimate,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$power,

    `power_0.50` = power.t.test(
      n = 10,
      delta = 0.50,
      sd = sd_estimate,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$power,

    `power_1.0` = power.t.test(
      n = 10,
      delta = 1.00,
      sd = sd_estimate,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$power,

    `power_2.0` = power.t.test(
      n = 10,
      delta = 2.00,
      sd = sd_estimate,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$power
  ) %>%
  ungroup() %>%

  pivot_longer(
    cols = starts_with("power_"),
    names_to = "log2delta",
    names_prefix = "power_",
    values_to = "power"
  ) %>%

  mutate(
    log2delta = as.numeric(log2delta),
    delta = 2^log2delta
    )
```

Based on the variability observed in our preliminary donors, 10 donors per group provides ≥80% power to detect differences in approximately the 1.4–2-fold range (depending on cell type and treatment). Dotted line at power=0.8 for visualization.

```{r fig.height=4, fig.width=4}
ggplot(
  power_results,
  aes(x = log2delta, y = power)
) +
  facet_grid(
    treatment~celltype,
    ) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  theme_bw() +
  theme(
    # aspect.ratio = 0.75,
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 10)
  ) +
  labs(
    x = "Δ expression (log2)",
    y = "power",
    title = "Power to detect differences in IRAK3\nexpression with n=10 per group"
  ) +
  scale_y_continuous(
    expand = expansion(mult = 0.2),
    breaks = c(0, 0.25, 0.5, 0.75, 1)
    )
```

Explicitly calculating the effect size detectable at 80% power, the proposed sample size of 10 normal and 10 severe asthma donors provides 80% power to detect approximately 1.2–1.6-fold differences in IRAK3 expression, depending on treatment and cell type.

```{r}
power_08 <- primary_tpm %>%
  filter(
    Gene == "IRAK3",
    celltype %in% c("ALI", "HBE")
  ) %>%
  
  group_by(treatment, celltype) %>%
  summarise(
    sd_estimate = sd(log2tpm),
    .groups = "drop"
  ) %>%
  
  rowwise() %>%
  mutate(
    
    power = 0.8,
    
    log2delta = power.t.test(
      n = 10,
      power = 0.8,
      sd = sd_estimate,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$delta,
    
    fold_change = 2^log2delta
    
  ) %>%
  ungroup()

kableExtra::kable(power_08)
```

```{r}
cleanup_objects()
```


